Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2019-03-15 RNARedPrint.tex
Changes for R2; leave some todo notes

@@

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, \revisedOne{which substantially reduce the need for costly wet-lab experiments.} \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the generation of sequences \revisedOne{that fold well into multiple target structures.} 
  % 
  \revisedOne{Based on constraint networks, our approach \ourprog{}} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  Moreover, we study general properties of our approach empirically and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  \revisedOne{established} design benchmark. \revisedOne{Our results demonstrate the efficacy and feasibility of the method in practice
  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}
%
  \revisedOne{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible, and effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

\revisedOne{As central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features. For the sake of clarity, we introduce this method specialized to the sampling of RNA sequences that have specific energies for multiple structures and specific \GCb\%.}

@@

%\paragraph{Central computational problem.}
Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revisedOne{$k\geq m$} features $F_1,\dots, F_k$\revisedOne{, typically representing the energies of the structures and additional sequence properties,} associated with weights $\pi_1,\dots,\pi_k$. Our goal is to sample sequences \revisedOne{$S$ (which satisfy the base pairing rules for all structures) from the Boltzmann distribution defined by} 
\begin{equation}
\label{eq:sample-distribution}
\Pr(S\mid\pi_1,\dots,\pi_k) \propto \displaystyle\prod_{1\leq \ell\leq k}\!\! \pi_\ell^{-F_\ell(S)} \end{equation}

The workhorse of our approach is the \revisedOne{fixed-parameter tractable} computation of feature-dependent partition functions over sequences, \revisedOne{namely partition functions of the form} 
%
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)},
  \end{equation}
  for specific weights $\pi_1,\dots,\pi_k$.

@@

$$
F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f\left(\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}\right).
$$
\revisedOne{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$; $p=|\dep(f)|$).}

@@

        Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions associated with local structural motifs (base pairs, base pair stacks, loops, \ldots) under a certain nucleotide assignment. Energy models can thus be captured generically by introducing, for each motif $m$ occurring in a target structure, a contribution $f_m$, taking a specific value for each assignment of nucleotides to its positions $\dep(f_m)$. For instance, the contribution of a \Def{base pair stack}, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
        We refer to energy models that consider the contributions of all base pair stacks (and thus introduce the corresponding dependencies) collectively as the \Def{stacking energy model}.
}

@@

\revisedOne{
In the simplest case, this set captures the requirement of canonical base pairing for each structure.
To express this, let us define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
and edges $\bigcup_{\ell\in[1,k]} R_\ell$. %Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.

@@

\smallskip
\noindent
The \Def{treewidth} of a tree decomposition $(T,\chi)$ is defined as
$\min_{u\in T} |\chi(u)| - 1.$ 
%The \Def{treewidth} of a graph $G$ is the smallest treewidth of any tree decomposition of $G$.
\end{definition}
%
\revisedOne{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependent sub-problems to ensure consistency in a recursive computation. Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}

@@

%For this purpose, we define the \Def{transformed contribution} 
%$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
%such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
%
Let us now specify the concrete set $\F$ of contributions that we use
for the design in the base pair model targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. The set $\F$ thus consists of
\begin{itemize}
\item the transformed contributions $\ln(\pi_0)f^{\GCb}_i$ for the ~$\GCb\%$ feature ($i=1,\dots,n$);
\item the transformed contributions $\ln(\pi_\ell)f^{R_\ell}_{ij}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
\end{itemize}

@@

We compute the partition function (as specified by $\F$) by dynamic programming based
on a tree decomposition of $G_{\F}$, the dependency graph associated with $\F$.
Note, that analogous algorithms could be easily derived to count valid sequences, or list sequences having minimal free-energy.
%This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.

@@

Since our algorithms iterate over specific sets of sequence positions, we moreover define
the \Def{set $\partseqs(\Y)$ of all partial sequences determining the positions of $\Y=\subseteq\{1,\dots,n\}$} in all combinations of nucleotides $\{\Ab,\Cb,\Gb,\Ub\}$,
i.e.~for $\Y=\{y_1,\dots,y_{q}\}$,
\begin{align*}
\partseqs(\Y) = \{ & \{ y_i\mapsto v_i \mid i=1,\dots,q \} \\& | (v_1,\dots,v_{q}) \in \{\Ab,\Cb,\Gb,\Ub\}^{q} \}.
\end{align*}
%
We assume the following properties of the given cluster tree (reflecting $\F$):
\begin{itemize}
\item
$T$ is
connected and contains a dedicated node $r$, with
$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$. If such a root does not exist, it can be added to the tree decomposition and connected to one node in each connected component of $T$;
\item
all edges in the tree decomposition are oriented towards this root;
\item
all sets $\difference{u}{v}$ are singleton: for any given
cluster tree, an equivalent (in term of treewidth) cluster tree can
always be obtained by inserting at most $\Theta(|\X|)$ additional
clusters.
\end{itemize}

Algorithm~\ref{alg:pf} computes the partition function by passing
messages along the directed edges $u\to v$ (which point from 
child $u$ to its parent $v$). Each message $m$ has the form of a contribution, i.e. it takes a partial sequence, depends
on the positions $\dep(m)\subseteq \X$, and yields a partition function
in $\real$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Supp.\ Mat.~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function (i.e.~due to proper encoding the partition function of our design problem) through
\begin{math}
  \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

%The complexity of the proposed sampling algorithm depends on the treewidth of the dependency graph $G_\F$.
% The treewidth $w$ of a cluster tree $(T,\chi,\phi)$, $T=(V,E)$, is $\max_{v\in V} |\chi(v)| - 1$, i.e.~the maximum number of positions in any of its clusters minus 1.
Let $s$ denote the maximum size of any separator set $\separator{u}{v}$ and $D$ denote the maximum size of
$\difference{u}{v}$ over $(u,v)\in E$.  In the absence
of specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and
$\mathcal{O}(|V|\cdot4^s)$ space; Alg.~\ref{alg:sampling} would
require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per sample on arbitrary
tree decompositions
(Supp.\ Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

@@

By this theorem, the complexity is polynomial for fixed value of $w$, and Boltzmann sampling in our setting is thus fixed parameter tractable (FPT) in the treewidth.
%
The complexity of the precomputation can be further improved to
$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ ($c\le w+1$) is the maximum number of connected components represented in a node of the tree decomposition (Supp.\ Mat.~\ref{appsec:improvedComplexity}).

Note that in this complexity analysis, we do not include time and space for computing the
tree decomposition itself, since we observed that the computation time
of tree decomposition (\Software{GreedyFillIn}, implemented in
\Software{LibTW}~by \citet{Dijk2006}) for multi-target sampling is
negligible compared to Alg.~\ref{alg:pf}
(Supp.\ Mat.~\ref{appsec:treedecomp} and
\ref{appsec:dependency-cliques}).

@@

\paragraph{Approximating Turner Energy using Simpler Energy Models.}
To capture the realistic Turner model $\EnergyTurner$ more efficiently, we exploit the tight correlation between
$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
(Supp.\ Mat.~\ref{appsec:modelparameters}). More precisely, we observed a
structure-specific affine dependency between the Turner and stacking
energy models, so that
$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
for any structure $R$ and sequence $S$. We inferred the
$(\gamma,\delta)$ parameters from a set of sequences generated with
homogeneous weights $w=e^{\beta}$, tuning only \GCb\% to a
predetermined value.  Finally, we adjusted the targeted energies
within our stacking model to
$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
order to reach, on average, the targeted energy
$\EnergyTurner^{\star}$ in the Turner model.

\subsection*{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
The flexibility of our framework allows to support the advanced sampling technique called "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to enforce (probabilistically)  additional, complex properties of the samples through an additional rejection.
This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and di-nucleotide content~\citep{Zhang2013} of sampled RNA sequences. Here, in addition to controlling \GCb\% (our feature $F_0$) we use it to target the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the individual target structures (features $F_1,\dots,F_m$).

@@

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution. In other words,  $\mathbb{E}(F_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$ and, analogously, the expectation of $F_0(S)$ is the targeted GC content.
The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp.\ Mat.~\ref{appsec:weight_derivatives}). More generally, computing a suitable parameter vector $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.

@@

For a discussion on the implications of our hardness results beyond exact counting see Supp.\ Mat.~\ref{appsec:approximate-counting}.
}

@@

\begin{figure}
	\includegraphics[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
	\caption{\revisedOne{Comparison of free-energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}
        \label{fig:BoltzmannBenefits}
\end{figure}

\revisedOne{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly. As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Supp. Mat. \ref{appsec:immediate-benefits-for-negative-design}, sequences generated in the Boltzmann distribution not only reach lower free-energies than those generated in a uniform setting, but also achieve better Boltzmann probabilities. While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor. The empirical superiority of Boltzmann sampling appears robust to the target structure length and topology, as demonstrated by prior work~\cite{Reinharz2013}.}

\begin{figure}
	  \begin{minipage}{.35\textwidth}\centering 
{\sffamily {\bfseries A} -- 
	Stacking model}\\
	\resizebox{\textwidth}{!}{
		\begin{tabular}{c|c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c}
		$\Delta G$ 	& \Ab\Ub & \Cb\Gb & \Gb\Cb & \Gb\Ub & \Ub\Ab & \Ub\Gb \\[.4em]\hline
			\Ab\Ub &
			-0.18&
			-1.13&
			-1.09&
			-0.38&
			-0.26&
			-0.62
			\\[.4em]
			\Cb\Gb&
			-1.11&
			-2.23&
			-1.89&
			-1.22&
			-1.10&
			-1.44
			\\[.4em]
			\Gb\Ub&
			-0.55&
			-1.26&
			-1.58&
			-0.72&
			-0.50&
			-0.69
			\\
	\end{tabular}}\\[3em]
{\sffamily {\bfseries C} -- Tree widths}\\
\includegraphics[width=\textwidth,trim=350 0 0 50,clip]{Figs/td-widths}
\end{minipage}
	  \begin{minipage}{.6\textwidth}\centering
	{\sffamily {\bfseries B} -- Correlation plot}\\
	  	\includegraphics[width=\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor} 
  	\end{minipage}

@@

\revisedOne{%
However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with treewidths scaling at least as large as the number of nucleotides in the largest loop. Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model. As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, whose design principles are further described in Supp. Mat.~\ref{appsec:modelparameters}, can be used to approximate the Turner energy model very adequately (correlation coefficient $R=0.99$) in the context of sequence design. Using this simpler model greatly reduces the treewidth, and thus the computational requirements of the whole method even for complex instances.
}

>>>>

\newcommand{\numTargetStructures}{\revised{k}}
\newcommand{\numFeatures}{\revised{m}}
\newcommand{\motif}{\revised{\frak m}}

@@

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, \revisedOne{which substantially reduce the need for costly wet-lab experiments.} \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the generation of sequences \revisedOne{that fold well into multiple target structures.} 
  % 
  \revisedOne{Based on constraint networks, our approach \ourprog{}} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  Moreover, we study general properties of our approach empirically and generate
  biologically relevant multi-target Boltzmann-weighted designs for an
  \revisedOne{established} design benchmark. \revisedOne{Our results demonstrate the efficacy and feasibility of the method in practice
  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}
%
  \revisedOne{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible, and effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

\revisedOne{As central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features. For the sake of clarity, we introduce this method specialized to the sampling of RNA sequences that have specific energies for multiple structures and specific \GCb\%.}
\revised{In this way, we address the positive design of RNA sequences. Positive design is contrasted to the often
desirable negative design of RNAs, which optimizes the stability of the target structures \emph{in relation to all
other potential structures}. Remarkably, the even more complex task of negative design immediately benefits from positive
design (Additional file 1: Section~\ref{appsec:immediate-benefits-for-negative-design}), which provides an initial motivation to study the positive design problem by itself.}

@@

%\paragraph{Central computational problem.}
Let us consider \revised{a set of} $\numTargetStructures$ (secondary) structures $\R = \{R_1,\dots, R_{\numTargetStructures}\}$, each abstracted as a set of base pairs, and
\revisedOne{$\numFeatures\geq \numTargetStructures$} features $F_1,\dots, F_{\numFeatures}$\revisedOne{, typically
representing the energies of the structures and additional sequence properties,} associated with weights
$\pi_1,\dots,\pi_{\numFeatures}$ \revised{in $\mathbb{R}^+$}. Our goal is to sample sequences \revisedOne{$S$ (which satisfy the base pairing rules for all structures) from the Boltzmann distribution defined by} 
\begin{equation}
\label{eq:sample-distribution}
\Pr(S\mid\pi_1,\dots,\pi_{\numFeatures}) \propto \displaystyle\prod_{1\leq \ell\leq \numFeatures}\!\! \pi_\ell^{-F_\ell(S)} \end{equation}

The workhorse of our approach is the \revisedOne{fixed-parameter tractable} computation of feature-dependent partition functions over sequences, \revisedOne{namely partition functions of the form} 
%
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_1,\dots,\pi_{\numFeatures}} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq \numFeatures} \pi_\ell^{-F_\ell(S)},
  \end{equation}
  for specific weights $\pi_1,\dots,\pi_{\numFeatures}$.

@@

$$
F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f\left(\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}\right).
$$
\revisedOne{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto
S_{x_p}}\right\}$ denotes the assignment, that assigns the respective nucleotides $S_{x_q}$ ($1\leq q\leq p$;
$p=|\dep(f)|$)} \revised{to the positions $x_q$  in $\dep(f)$}.

@@

Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions
associated with local structural motifs (base pairs, base pair stacks, loops, \ldots) under a certain nucleotide
assignment. Energy models can thus be captured generically by introducing, for each motif $\motif$ occurring in
a target structure, a contribution $f_{\motif}$, taking a specific value for each assignment of nucleotides to its
positions $\dep(f_{\motif})$. For instance, the contribution of a \Def{base pair stack}, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
We refer to energy models that consider the contributions of all base pair stacks (and thus introduce the corresponding dependencies) collectively as the \Def{stacking energy model}.
}

@@

\revisedOne{
In the simplest case, this set captures the requirement of canonical base pairing for each structure.
To express this, let us define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
and edges $\bigcup_{\ell\in[1,\numTargetStructures]} R_\ell$.

@@

\smallskip
\noindent
The \Def{treewidth} of a tree decomposition $(T,\chi)$ is defined as
$\max_{u\in T} |\chi(u)| - 1.$ 
%The \Def{treewidth} of a graph $G$ is the smallest treewidth of any tree decomposition of $G$.
\end{definition}
%
\revisedOne{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependent sub-problems to ensure consistency in a recursive computation. Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}

@@

%For this purpose, we define the \Def{transformed contribution} 
%$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
%such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
%
Let us now specify the concrete set $\F$ of contributions that we use
for the design in the base pair model\todo{Didn't we plan to get rid of the bp model?}{} targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. The set $\F$ thus consists of
\begin{itemize}
\item the transformed contributions $\ln(\pi_0)f^{\GCb}_i$ for the ~$\GCb\%$ feature ($i=1,\dots,n$);
\item the transformed contributions $\ln(\pi_\ell)f^{R_\ell}_{ij}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
\end{itemize}

@@

We compute the partition function (as specified by $\F$) by dynamic programming based
on a tree decomposition of $G_{\F}$, the dependency graph associated with $\F$.
Note, that analogous algorithms could be easily derived to count valid sequences, or list sequences having minimum free
energy.
%This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.

@@

Since our algorithms iterate over specific sets of sequence positions, we moreover define
the \Def{set $\partseqs(\Y)$ of all partial sequences determining the positions of $\Y\subseteq\{1,\dots,n\}$} in all combinations of nucleotides $\{\Ab,\Cb,\Gb,\Ub\}$,
i.e.~for $\Y=\{y_1,\dots,y_{q}\}$,
\begin{align*}
\partseqs(\Y) = \{ & \{ y_i\mapsto v_i \mid i=1,\dots,q \} \\& | (v_1,\dots,v_{q}) \in \{\Ab,\Cb,\Gb,\Ub\}^{q} \}.
\end{align*}
%
We assume the following properties of the given cluster tree (reflecting $\F$):
\begin{itemize}
\item
$T$ is
connected and contains a dedicated node $r$, with
$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$. If such a root does not exist, it can be added to the tree decomposition and connected to one node in each connected component of $T$;
\item
all edges in the tree decomposition are oriented towards this root;
\item
all sets $\difference{u}{v}$ are singleton: for any given
cluster tree, an equivalent (in term of treewidth) cluster tree can
always be obtained by inserting at most $\Theta(|\X|)$ additional
clusters.
\end{itemize}

Algorithm~\ref{alg:pf} computes the partition function by passing
messages along the directed edges $u\to v$ (which point from 
child $u$ to its parent $v$). Each message $m$ has the form of a contribution, i.e. it takes a partial sequence, depends
on the positions $\dep(m)\subseteq \X$, and yields a partition function
in $\real$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Additional file~1: Section~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function (i.e.~due to proper encoding the partition function of our design problem) through
\begin{math}
  \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

%The complexity of the proposed sampling algorithm depends on the treewidth of the dependency graph $G_\F$.
% The treewidth $w$ of a cluster tree $(T,\chi,\phi)$, $T=(V,E)$, is $\max_{v\in V} |\chi(v)| - 1$, i.e.~the maximum number of positions in any of its clusters minus 1.
Let $s$ denote the maximum size of any separator set $\separator{u}{v}$ and $D$ denote the maximum size of
$\difference{u}{v}$ over $(u,v)\in E$.  In the absence
of specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and
$\mathcal{O}(|V|\cdot4^s)$ space; Alg.~\ref{alg:sampling} would
require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per sample on arbitrary
tree decompositions
(Additional file 1: Section~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

@@

By this theorem, the complexity is polynomial for fixed value of $w$, and Boltzmann sampling in our setting is thus fixed parameter tractable (FPT) in the treewidth.
%
The complexity of the precomputation can be further improved to
$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ ($c\le w+1$) is the maximum number of connected components represented in a node of the tree decomposition (Additional file~1: Section~\ref{appsec:improvedComplexity}).

Note that in this complexity analysis, we do not include time and space for computing the
tree decomposition itself, since we observed that the computation time
of tree decomposition (\Software{GreedyFillIn}, implemented in
\Software{LibTW}~by \citet{Dijk2006}) for multi-target sampling is
negligible compared to Alg.~\ref{alg:pf}
(Additional file~1: Sections~\ref{appsec:treedecomp} and
\ref{appsec:dependency-cliques}).

@@

\paragraph{Approximating Turner Energy using Simpler Energy Models.}
To capture the realistic Turner model $\EnergyTurner$ more efficiently, we exploit the tight correlation between
$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
(Additional file~1: Section~\ref{appsec:modelparameters}). More precisely, we observed a
structure-specific affine dependency between the Turner and stacking
energy models, so that
$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
for any structure $R$ and sequence $S$. We inferred the
$(\gamma,\delta)$ parameters from a set of sequences generated with
homogeneous weights $w=e^{\beta}$, tuning only \GCb\% to a
predetermined value.  Finally, we adjusted the targeted energies
within our stacking model to
$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
order to reach, on average, the targeted energy
$\EnergyTurner^{\star}$ in the Turner model.

\subsection*{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
The flexibility of our framework allows to support the advanced sampling technique called "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to enforce (probabilistically)  additional, complex properties of the samples through an additional rejection.
This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and di-nucleotide
content~\citep{Zhang2013} of sampled RNA sequences. Here, in addition to controlling \GCb\% (our feature $F_0$) we use
it to target the free energies $(\TargetE_1,\ldots,\TargetE_{\numTargetStructures})$ of the individual target structures (features
$F_1,\dots,F_{\numFeatures}$).

@@

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution. In other words,  $\mathbb{E}(F_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$ and, analogously, the expectation of $F_0(S)$ is the targeted GC content.
The expected value of $F_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Additional file~1: Section~\ref{appsec:weight_derivatives}).
More generally, computing a suitable parameter vector $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.

@@

For a discussion on the implications of our hardness results beyond exact counting see Additional file~1: Section~\ref{appsec:approximate-counting}.
}

@@

\begin{figure}
	\includegraphics[width=0.8\textwidth]{Figs/Plots/doublestem_weight1-500_mean}
	\caption{\revisedOne{Comparison of free energy and Boltzmann probability for 10~000 uniform ($\pi=1$; blue dots) and Boltzmann distributed ($\pi=500$; green dots) sequences, targeting a simple structure consisting of two adjacent helices.}}
        \label{fig:BoltzmannBenefits}
\end{figure}

\revisedOne{First, we show how seed sequences can be generated in a Boltzmann distribution, leading to designs that are substantially more stable that those generated uniformly. As can be seen in Figure~\ref{fig:BoltzmannBenefits} and Additional file~1: Section~\ref{appsec:immediate-benefits-for-negative-design}, sequences generated in the Boltzmann distribution not only reach lower free-energies than those generated in a uniform setting, but also achieve better Boltzmann probabilities. While the former is expected since the Boltzmann distribution explicitly favors low-energy candidates, the latter is somewhat surprising, since the Boltzmann probability of a target structure could, in principle, decrease under Boltzmann sampling due to the partition function growing faster than the Boltzmann factor. The empirical superiority of Boltzmann sampling appears robust to the target structure length and topology, as demonstrated by prior work~\cite{Reinharz2013}.}

\begin{figure}
	  \begin{minipage}{.35\textwidth}\centering 
{\sffamily {\bfseries A} -- 
	Stacking model}\\
	\resizebox{\textwidth}{!}{
		\begin{tabular}{c|c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c@{\hspace{.5em}}c}
		$\Delta G$ 	& \Ab\Ub & \Cb\Gb & \Gb\Cb & \Gb\Ub & \Ub\Ab & \Ub\Gb \\[.4em]\hline
			\Ab\Ub &
			-0.18&
			-1.13&
			-1.09&
			-0.38&
			-0.26&
			-0.62
			\\[.4em]
			\Cb\Gb&
			-1.11&
			-2.23&
			-1.89&
			-1.22&
			-1.10&
			-1.44
			\\[.4em]
			\Gb\Ub&
			-0.55&
			-1.26&
			-1.58&
			-0.72&
			-0.50&
			-0.69
			\\
	\end{tabular}}\\[3em]
{\sffamily {\bfseries C} -- Treewidths}\\
\includegraphics[width=\textwidth,trim=350 0 0 50,clip]{Figs/td-widths}
\end{minipage}
	  \begin{minipage}{.6\textwidth}\centering
	{\sffamily {\bfseries B} -- Correlation plot}\\
	  	\includegraphics[width=\textwidth,trim=0 0 0 50,clip]{Figs/stackingcor} 
  	\end{minipage}

@@

\revisedOne{%
However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with treewidths scaling at least as large as the number of nucleotides in the largest loop. Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model. As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, whose design principles are further described in Additional file~1: Section~\ref{appsec:modelparameters}, can be used to approximate the Turner energy model very adequately (correlation coefficient $R=0.99$) in the context of sequence design. Using this simpler model greatly reduces the treewidth, and thus the computational requirements of the whole method even for complex instances.
}
